Natural selection in the Health and Retirement Study

### 2023-11-22 hrs-selection.qmd
TE changes adding PGI repository scores and grandkid regression

---
title: "Natural selection in the Health and Retirement Study"
author: "David Hugh-Jones"
abstract: |
  I investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for
  natural selection: among both white and black respondents,
  scores which correlate negatively (positively) with education are
  selected for (against). Selection coefficients are larger among low-income
  and unmarried parents, but not among younger parents or those with less 
  education. I also estimate effect sizes corrected for noise in the 
  polygenic scores. 
date: "September 2023"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

# TODO: ratio to family poverty threshold, h10inpovr
# TODO: I think for all these guys, the raevborn question will have been 
#.   asked by the time they were indeed over 45/40. Double check.

# Is ADHD reverse-coded? - Doesn't seem to be, correlates negatively
#       with income
# What about XDISORDER?
# ADHD doesn't seem to be. Nor does XDISORDER (correlates with e.g. r*depres)

options(digits = 3)
set.seed(271075)

@@

nice_names <- c(
  SCZ         = "Schizophrenia",
  WC          = "Waist circ.",
  WHR         = "Waist-hip ratio",
  NEUROT      = "Neuroticism",
  WELLB       = "Well-being",
  DEPSYMP     = "Depr. symptoms",
  CD          = "Cor. art. disease",
  MI          = "Myoc. infarction",
  CORT        = "Cortisol",
  T2D         = "Type 2 diabetes",
  BIP         = "Bipolar",
  ADHD_PGC17  = "ADHD",
  XDISORDER   = "Cross-disorder",
  MENA        = "Age at menarche",
  MENO        = "Age at menopause",
  EXTRAVER    = "Extraversion",
  AUTISM      = "Autism",
  AB          = "Antisoc. behav.",
  OCD         = "OCD",
  AFBC        = "Age 1st birth combn",
  AFBF        = "Age 1st birth F",
  AFBM        = "Age 1st birth M",
  MDD2        = "Maj. depr. disorder",
  PTSDAA      = "PTSD Afr",
  PTSDEA      = "PTSD Eur",
  PTSDC       = "PTSD combined",
  HDL         = "HDL cholesterol",
  LDL         = "LDL cholesterol",
  TC          = "Total cholesterol",
  ANXFS       = "Anxiety factor", 
  ANXCC       = "Anxiety case-control",
  BUN         = "Blood urea nitro.",
  BUNTE       = "Blood urea nitro. TE",
  CKD         = "Kidney disease",
  CKDTE       = "Kidney disease TE",
  DBP         = "Diastolic blood pr.",
  BMI2        = "BMI",
  HEIGHT2     = "Height",
  AI          = "Age smoking init.",
  CPD_GSCAN19 = "Cigs per day",
  DPW         = "Drinks per week",
  SC          = "Smoking cessation",
  SI          = "Smoking initiation",
  HTN         = "Hypertension",
  CANNABIS    = "Cannabis",
  GWAD2NA     = "Alzheimer's no APOE",
  ALZ_01AD2NA = "Alzheimer's p 0.01 no APOE",
  GWAD2WA     = "Alzheimer's",
  ALZ_01AD2WA = "Alzheimer's p 0.01",
  ALC         = "Alcohol dependence",
  PP          = "Pulse pressure",
  SBP         = "Systolic blood pr.",
  EGFR        = "Glom. filtr.",
  EGFRTE      = "Glom. filtr. TE",
  EA3         = "Educ. attainment",
  HBA1CAA     = "HbA1c Afr",
  HBA1CEA     = "HbA1c Eur",
  GCOG2       = "Gen. cognition"
)

@@

@@

old_pgs <- c("EDU2", "EDU3", "HEIGHT", "BMI",
           "AD", "AD2", "GENCOG", "EVRSMK", "LONG", "CPD_TAG10",
           "ADHD_PGC10", "MDD")
pgs <- pgen4e |>
    rename_with(shorten) |> 
    # weird that I have to operate on whole df to use select semantics
    select(-hhid, -pn, -version,
           -matches("^PC\\d"),
           -matches("^NEB"), # number ever born
           # outdated, or EDU3 without 23andMe:
           -any_of(old_pgs)
           ) |>
    names()

# weighting
rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
rand$weight <- ifelse(is.na(rand$MBIOWGTR), rand$NBIOWGTR, rand$MBIOWGTR)
# compensate for fecund parents more likely to have child in survey.
# we use living sibs rather than including dead sibs too here. I think that's
# the correct approach:
rand$parent_weight <- rand$weight / (rand$r10livsib + 1)

@@

rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_raevbrn = mean(raevbrn, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )

rand$neurot <- rand |> 
  select(matches("r\\d+lbneur")) |> 
  rowMeans(na.rm = TRUE)

# two of four Alzheimer's variants with awkward names:
pgs <- str_replace(pgs, "^01", "ALZ_01")
names(rand) <- str_replace(names(rand), "^01", "ALZ_01")

@@

# clean up cache so it doesn't get too big
rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(old_pgs), 
         all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
         rabyear, born, raehsamp, raestrat, parent_weight, ragender,
         adhd_score, neurot, r10diabe, r10hearte, r10cesd, age_smoked,
         anxiety, r10cogtot)

@@

@@

@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from @becker1976child. The theory has two components.

@@

Here, I make an independent test of the theory in the US population, using the Health and Retirement Survey [@hrsrandfam2023; @hrsrand2023]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The HRS also provides precalculated polygenic scores [@hrspgs2020] for both black and (non-hispanic) white participants, so I can check whether the theory predicts patterns of natural selection in both these ethnicities.

@@

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. I include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.

Genotyping took place in 2006, 2008 and subsequent years. I discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. I also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs)` scores. PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, I control for ten within-ethnicity principal components of the DNA array data.

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by mean number of children of people born in the same year. RLRS is calculated pooling ethnicities, i.e. treating them as members of the same biological population.

The HRS contains weights which match survey respondents to the US population. I use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, I weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012. Statistical tests are adjusted for clustering and stratification using the R "survey" package [@lumleysurvey2023].

@@

tidy_pgs <- mod_pgs <- list()
tidy_pgs_ed <- mod_pgs_edyrs <- list()
tidy_pgs_sibs <- mod_pgs_sibs <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
tidy_pgs_x_agefb <- mod_pgs_x_agefb <- list()

coefs_fert_edu <- cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
tidy_fert_edu_phen <- list()

for (eth in eths) {
  r_eth <- rand |> subset(ethnicity == eth)
  r_parents_eth <- r_parents |> subset(ethnicity == eth)

@@

@@

@@

@@

I estimated coefficients of PGS on RLRS among black and white respondents separately. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix @fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample. Standard errors are large because of the relatively low sample sizes, and only `r n_sig_bonf` score is significant at Bonferroni-corrected p \< 0.05/`r n_tests`. But I am most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

cors_fert_edu <- cors_sibs_edu <- list()
for (eth in eths) {
  rand_boot <- rand |> 
    subset(ethnicity == eth) |> 
    as.svrepdesign(type = "bootstrap", replicates = n_reps)
  r_parents_boot <- r_parents |> 
    subset(ethnicity == eth) |> 
    as.svrepdesign(type = "bootstrap", replicates = n_reps)

  cors_fert_edu[[eth]] <- 
    withReplicates(rand_boot, theta = calc_cors_edu, var = "rlrs")
  cors_sibs_edu[[eth]] <- 
    withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
}

@@

ci_boot_fert <- map(cors_fert_edu, \(x) confint(x)[1,])
ci_boot_sibs <- map(cors_sibs_edu, \(x) confint(x)[1,])

ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint(x)[2,])
ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint(x)[2,])

>>>>

---
title: "Natural selection in the Health and Retirement Study"
author: 
  - David Hugh-Jones
  - Tobias Edwards
abstract: |
  We investigate natural selection on polygenic scores
  in the contemporary US, using the Health and Retirement Study. Results
  partially support the economic theory of fertility as an explanation for
  scores which correlate negatively (positively) with education are
  selected for (against). Selection coefficients are larger among low-income
  and unmarried parents, but not among younger parents or those with less 
  education. We also estimate effect sizes corrected for noise in the 
  polygenic scores. 
  natural selection: among both white and black respondents,
date: "November 2023"
format: 
  pdf:
    keep-tex: true
    include-in-header:
      - text: |
          \usepackage{placeins}
    geometry:
    - top=2in
    - left=1.25in
    - right=1.25in
    - heightrounded
editor: visual
editor_options: 
  chunk_output_type: console
knitr:
  opts_chunk: 
    echo: false
bibliography: bibliography-hrs.bib
---

@@

# TE quick TODOs change reference to time being insignificant. Fix the footnote citations
options(digits = 3)
set.seed(271075)

@@

nice_names <- c(
  SCZ         = "Schizophrenia",
  WC          = "Waist circ.",
  WHR         = "Waist-hip ratio",
  NEUROT      = "Neuroticism",
  WELLB       = "Well-being",
  DEPSYMP     = "Depr. symptoms",
  CD          = "Cor. art. disease",
  MI          = "Myoc. infarction",
  CORT        = "Cortisol",
  T2D         = "Type 2 diabetes",
  BIP         = "Bipolar",
  ADHD_PGC17  = "ADHD",
  XDISORDER   = "Cross-disorder",
  MENA        = "Age at menarche",
  MENO        = "Age at menopause",
  EXTRAVER    = "Extraversion",
  AUTISM      = "Autism",
  AB          = "Antisoc. behav.",
  OCD         = "OCD",
  AFBC        = "Age 1st birth combn",
  AFBF        = "Age 1st birth F",
  AFBM        = "Age 1st birth M",
  MDD2        = "Maj. depr. disorder",
  PTSDAA      = "PTSD Afr",
  PTSDEA      = "PTSD Eur",
  PTSDC       = "PTSD combined",
  HDL         = "HDL cholesterol",
  LDL         = "LDL cholesterol",
  TC          = "Total cholesterol",
  ANXFS       = "Anxiety factor", 
  ANXCC       = "Anxiety case-control",
  BUN         = "Blood urea nitro.",
  BUNTE       = "Blood urea nitro. TE",
  CKD         = "Kidney disease",
  CKDTE       = "Kidney disease TE",
  DBP         = "Diastolic blood pr.",
  BMI2        = "BMI",
  HEIGHT2     = "Height",
  AI          = "Age smoking init.",
  CPD_GSCAN19 = "Cigs per day",
  DPW         = "Drinks per week",
  SC          = "Smoking cessation",
  SI          = "Smoking initiation",
  HTN         = "Hypertension",
  CANNABIS    = "Cannabis",
  GWAD2NA     = "Alzheimer's no APOE",
  ALZ_01AD2NA = "Alzheimer's p 0.01 no APOE",
  GWAD2WA     = "Alzheimer's",
  ALZ_01AD2WA = "Alzheimer's p 0.01",
  ALC         = "Alcohol dependence",
  PP          = "Pulse pressure",
  SBP         = "Systolic blood pr.",
  EGFR        = "Glom. filtr.",
  EGFRTE      = "Glom. filtr. TE",
  EA3         = "Educ. attainment",
  HBA1CAA     = "HbA1c Afr",
  HBA1CEA     = "HbA1c Eur",
  GCOG2       = "Gen. cognition",
  ACTIVITY="Physical Activity", #pgirep here
  ADHD="ADHD",
  ADVENTURE="Adventurous",
  AFB="Age First Birth",
  ALLERGYCAT="Cat Allergy",
  ALLERGYDUST="Dust Allergy",
  ALLERGYPOLLEN="Pollen Allergy",
  ASTECZRHI="Asthma/Eczema/Rhinitis",
  ASTHMA="Asthma",
  AUDIT="Alcohol Misuse",
  BMI="Body Mass Index",
  CANNABISpgirep="Cannabis Use",
  COGEMP="Cognitive Empathy",
  COPD="Chronic Obstructive Pulmonary Disease",
  CPD="Cigarettes per Day",
  CP="Cognitive Performance",
  DELAYDISC="Delay Discounting",
  DEP="Depressive Symptoms",
  DPWpgirep="Drinks per Week",
  EA="Educational Attainment",
  EVERSMOKE="Ever Smoker",
  EXTRA="Exraversion",
  FAMSAT="Satisfaction with Family",
  FINSAT="Satisfaction with Finances",
  FRIENDSAT="Satisfaction with Friends",
  HAYFEVER="Hayfever",
  HEIGHTpgirep="Height",
  HIGHMATH="Highest Math",
  LEFTOUT="Left Out of Social Activity",
  LONELY="Loneliness",
  MENARCHE="Age First Menses",
  MIGRAINE="Migraine",
  MORNING="Morning Person",
  NARCIS="Narcissism",
  NEARSIGHTED="Nearsightedness",
  NEBmen="Number of Children (men)",
  NEBwomen="Number of Children (women)",
  NEURO="Neuroticism",
  OPEN="Openness",
  READING="Age Started Reading",
  RELIGATT="Religious Attendance",
  RISK="Risk Tolerance",
  SELFHEALTH="Self-Rated Health",
  SELFMATH="Self-Rated Math Ability",
  SWB="Subjective Well-Being",
  VOICEDEEP="Age Voice Deepened",
  WORKSAT="Satisfaction with Work"
)

@@

# TE edit PGI repository EUR only
pgirep <- readr::read_tsv("data/pgi-repository/HRS_PGIrepo_v1.0.txt", show_col_types = FALSE)
names(pgirep)[1:2] <- c("hhid","pn")

# TE Maybe not the most efficient method, but this code removes MTAG PGS
pgirep <- select(pgirep,"hhid","pn",matches("PC"),matches("single"))

# TE Going to remove the NEB PGS here
pgirep <- select(pgirep, -c(PGI_NEBwomen_single) )

# TE this creates simple names for PGS in PGI Rep
pgi_rep_names <- str_split_fixed(names(pgirep)[23:57],"_",n=3)[,2]
names(pgirep)[23:57] <- pgi_rep_names

# TE PGI rep does not come pre-standardised. Going to standardize over whole sample to match
# the approach taken with PGENscores
pgirep[,3:ncol(pgirep)] <- scale(pgirep[,3:ncol(pgirep)])

@@

old_pgs <- c("EDU2", "EDU3", "HEIGHT", "BMI",
           "AD", "AD2", "GENCOG", "EVRSMK", "LONG", "CPD_TAG10",
           "ADHD_PGC10", "MDD")

# TE going to simplify names of pgen4e. Then merge, then create pgs.
pgen4e_new_names <- pgen4e |>
    rename_with(shorten) |> 
    # weird that we have to operate on whole df to use select semantics
    select(-version,
           -matches("^PC\\d"),
           -matches("^NEB"), # number ever born
           # outdated, or EDU3 without 23andMe:
           -any_of(old_pgs)
           )

pgirep <- select(pgirep, -matches("^PC\\d"))
#|>
#|
pgen4e_new_names <- left_join(pgen4e_new_names, pgirep, by = join_by(hhid == hhid, pn == pn),suffix=c("","pgirep"))

# TE Height is a duplicated name, but it's with the old PGS, so fixing this manually
colnames(pgen4e_new_names)[which(names(pgen4e_new_names) == "HEIGHT")] <- "HEIGHTpgirep"

# TE here we removed pgenscores that are duplicated in the pgi repository
# we also remove trans ethnic and African ancestry PGS
dupe_PGS <- c("EA3", # TE when I've tested it the pgirep one is slightly better...
              "GCOG2", #CP
              "BMI2",
              "NEUROT",
              "WELLB",
              "DEPSYMP",
              "ADHD_PGC17",
              "EXTRAVER",
              "AFBC", 
              "PTSDAA", # African    
              "PTSDC", # Combined Euro + African.
              #"ANXFS"  ,"ANXCC", continuous vs case-control anxiety 
              # r = 0.6 let's keep them
              # we suppose it's equivalent to having CP and EA
              "BUNTE", #trans ethnic BUNTE is kept 
              "CKDTE", 
              "BMI2",
              "HEIGHT2",
              "CPD_GSCAN19",
              "DPW",
              "SI",
              "CANNABIS", 
              "GWAD2NA", 
              "ALZ_01AD2NA",
              "GWAD2WA", # Three alzheimers deleted. Ones using p value 1 threshold
              # and alzheimers that remove APOE variant region
              "ALC", #AUDIT is alcohol misuse in pgirep
 	           "EGFR",
	          "HBA1CEA" #African ancestry vs European ancestry. No TE
              )

dupe_PGS_black <- c(
              "PTSDAA", # African    
              "PTSDC", # Combined Euro + African.
              #"ANXFS"  ,"ANXCC", continuous vs case-control anxiety 
              # r = 0.6 let's keep them
              # we suppose it's equivalent to having CP and EA
              "BUNTE", #trans ethnic BUNTE is kept 
              "CKDTE", 
              "BMI2",
              "CANNABIS", 
              "GWAD2NA", 
              "ALZ_01AD2NA",
              "GWAD2WA", # Three alzheimers deleted. Ones using p value 1 threshold
              # and alzheimers that remove APOE variant region
 	           "EGFR",
	          "HBA1CEA" #African ancestry vs European ancestry. No TE
              )

pgen4e_new_names <- select(pgen4e_new_names,-hhid,-pn,-any_of(dupe_PGS),-any_of(old_pgs))
pgs_white <- names(pgen4e_new_names)
pgs_white[which(pgs_white=="HEIGHT")] <- "HEIGHTpgirep"

pgs_black <- pgen4a |>
    rename_with(shorten) |> 
    select(-hhid, -pn, -version,
           -matches("^PC\\d"),
           -matches("^NEB"), # number ever born
           # outdated, or EDU3 without 23andMe:
           -any_of(old_pgs),
           -any_of(dupe_PGS_black)
           ) |>
    names()

rand <- left_join(rand, pgirep, by = join_by(hhid, pn),suffix=c("","pgirep"))

# weighting
rand <- left_join(rand, trk_weights, by = join_by(hhid == HHID, pn == PN))
rand$weight <- ifelse(is.na(rand$MBIOWGTR), rand$NBIOWGTR, rand$MBIOWGTR)

# TE I'm replacing 0 weights with NA because it can create odd problems
rand$weight <- na_if(rand$weight, 0)

# compensate for fecund parents more likely to have child in survey.
# we use living sibs rather than including dead sibs too here. we think that's
# the correct approach:
rand$parent_weight <- rand$weight / (rand$r10livsib + 1)
rand$child_weight <- rand$weight*rand$raevbrn

@@

# TE this is average number of kids that your kids have
rand$div_h10gkid <- rand$h10gkid/rand$raevbrn

# TE 283 people report having grandkids, but having had no children. Weird!
gkids_unusable_white <- sum(na.omit(ifelse(subset(rand, ethnicity == "white")$h10gkid > 0 & subset(rand, ethnicity == "white")$raevbrn==0,1,0)))

gkids_unusable_black <- sum(na.omit(ifelse(subset(rand, ethnicity == "black")$h10gkid > 0 & subset(rand, ethnicity == "black")$raevbrn==0,1,0)))

# If grandkids but no kids div_gkid is infinite. Replacing with NA
rand$div_h10gkid[is.infinite(rand$div_h10gkid)] <- NA

# TE added some variables for grandkid analysis
# TE I've also added sampling weights in calculation of mean values
# note the child weight is just weight* number of kids. so mean_h10gkid 
# and mean_div_h10gkid are actually the same
rand <- rand |>
  mutate(.by = rabyear, # calculate this pooling ethnicity
    mean_weight = mean(weight, na.rm = TRUE),
    mean_child_weight = mean(child_weight, na.rm = TRUE),
    mean_raevbrn = mean(raevbrn*weight/mean_weight, na.rm = TRUE),
    rlrs = raevbrn/mean_raevbrn,
    mean_h10gkid = mean(h10gkid*weight/mean_weight, na.rm = TRUE),
    rlrs_h10gkid = h10gkid/mean_h10gkid,
    mean_div_h10gkid = mean(div_h10gkid*child_weight/mean_child_weight, na.rm = TRUE), 
    rlrs_div_h10gkid = div_h10gkid/mean_div_h10gkid
  ) |>
  mutate(.by = ethnicity, # calculate these stats within-ethnicity
    income_resid = calc_income(rand, ethnicity[1]),
    income_resid = income_resid - min(income_resid, na.rm = TRUE) + 1,
    income_med = chop_equally(income_resid, 2, labels = c("Low", "High")),
    # as.numeric avoids haven.labelled issues:
    raedyrs = as.numeric(raedyrs),
    r10hearte = as.numeric(r10hearte),
    r10diabe = as.numeric(r10diabe),
    edu = chop(as.numeric(raedyrs), 13, lbl_discrete(symbol = "-")),
    born = chop(rabyear, 1942, labels = c("pre-1942", "post-1942")),
    # wave 10 is 2010
    # table(rands$white$r10mstat)
     # 1                 1.married
     # 2   2.married,spouse absent
     # 3               3.partnered
     # 4               4.separated
     # 5                5.divorced
     # 6      6.separated/divorced
     # 7                 7.widowed
     # 8           8.never married
    married = factor(r10mstat == 1, 
                     levels = c(FALSE, TRUE),
                     labels = c("Other", "Married")),
    r10sibs = r10livsib + 1, # n. sibs including oneself!
    agefbn = r10agey_e - h10ageokid,
    agefbn = ifelse(agefbn < 12, NA_real_, agefbn),
  ) |> 
  mutate(.by = c(ethnicity, ragender),
    agefb = chop_equally(agefbn, 2, labels = c("Low", "High"))
  )
# TE I've plotted the mean values by year. They look reasonably stable. Beauchamp 2016 took a three year average. we don't think this is necessary

rand$neurot <- rand |> 
  select(matches("r\\d+lbneur")) |> 
  rowMeans(na.rm = TRUE)
# two of four Alzheimer's variants with awkward names:
#pgs <- str_replace(pgs, "^01", "ALZ_01")
pgs_white <- str_replace(pgs_white, "^01", "ALZ_01")
pgs_black <- str_replace(pgs_black, "^01", "ALZ_01")
names(rand) <- str_replace(names(rand), "^01", "ALZ_01")
pgs <- unique(c(pgs_white,pgs_black))

# Child dataset testing
# TE sample size not too small... But mum and dad often report different values of grandkids
# and big questions of sampling bias. Given the weakness of the data I don't think the approach can be justified

#rand_male <- subset(rand,ragender == 1)
#rand_female <- subset(rand,ragender == 2)

#rand_mid <- merge(rand_male,rand_female, by = "hhid",suffixes=c(".m",".f"),all.x=FALSE,all.y=FALSE)
#rand_mid <- subset(rand_mid,ethnicity.m=="white"))
# 3588

#rand_mid$EA <- rand_mid$EA.m + rand_mid$EA.f
#install.packages("fixest")
#library(fixest)

#cor(rand_mid$EA3.f,rand_mid$EA3.m)
#reg <- lm(scale(EA3.f) ~ scale(EA3.m), data=rand_mid)
#reg <- feols(scale(EA3.f) ~ scale(EA3.m), data=rand_mid)
#reg <- feols(rlrs.m ~ EA3.m, data=rand_mid)
#rand_mid$gkid <- rand_mid$h10gkid.m/rand_mid$raevbrn.m
#rand_mid$gkidf <- rand_mid$h10gkid.f/rand_mid$raevbrn.f

#reg <- feols(gkid ~ EA, data=rand_mid, ~ child_weight.m)

#reg <- feols(gkidf ~ EA, data=rand_mid, ~ child_weight.f)

#reg <- feols(h10gkid.m ~ EA, data=rand_mid, ~ child_weight.m)
#reg <- feols(h10gkid.f ~ EA, data=rand_mid, ~ child_weight.f)

@@

# clean up cache so it doesn't get too big
rm(rand_orig, trk_orig, rands, rand_fam, rand_2010, rand_2016)
# TE we can delete more
rm(pgen,pgen4a,pgen4e,pgirep,trk_weights)

rand <- rand |> 
  select(ethnicity, all_of(pgs), all_of(old_pgs), 
         all_of(pcs), weight, raevbrn, rlrs,
         raedyrs, r10sibs, edu, income_med, married, agefb, agefbn,
         rabyear, born, raehsamp, raestrat, parent_weight, ragender,
         adhd_score, neurot, r10diabe, r10hearte, r10cesd, age_smoked,
         anxiety, r10cogtot, rlrs_div_h10gkid, child_weight)
# added my grandkid rlrs

@@

r_child <- svydesign(
    id = ~ raehsamp,
		strata = ~ raestrat,
		weights = ~ child_weight, 
		nest = TRUE,
		data = rand |> drop_na(child_weight) 
		)

@@

@hugh2022human explain patterns of natural selection on polygenic scores in the UK, using an economic theory of fertility derived from \@becker1976child. The theory has two components.

@@

Here, we make an independent test of the theory in the US population, using the Health and Retirement Survey \[\@hrsrandfam2023; \@hrsrand2023\]. The HRS is more representative of the population than UK Biobank, which addresses one potential weakness of the previous paper. The HRS also provides precalculated polygenic scores \[\@hrspgs2020\] for both black and (non-hispanic) white participants, so we can check whether the theory predicts patterns of natural selection in both these ethnicities.

@@

The HRS sample focuses on cohorts born between 1920 and 1960, but contains some younger and older participants. we include only male participants born before 1965 and female participants born before 1970, which guarantees that most will have completed their fertility by 2010. The resulting sample contains `r n_white` genotyped white participants and `r n_black` genotyped black participants.

<!--#  TE sources of pgs are not mentioned in the data section. I've added that since we are now using two sources -->

Genotyping took place in 2006, 2008 and subsequent years. PGS were taken from those pre-calculated by the HRS [@hrspgs2020] and those produced by the Social Science Genetic Association Consortium, as part of their Polygenic Index Repository [@becker2021resource]. Scores created by the HRS were provided for black and white participants, but Polygenic Index Repository scores were only created for white participants.

Multiple steps were taken to prune the polygenic scores. For the white participants, when scores from the two samples measured the same trait, we only used the PGS from the Polygenic Index Repository. For some traits, polygenic scores were created from European ancestry GWAS and GWAS with samples of different ancestry. We choose to use polygenic scores trained only on individuals of European ancestry. we discard obsolete PGS for which there is a newer, more accurate score targeting the same phenotype. we also discard PGS for number of children ever born (but keep scores for age at first birth). This leaves a total of `r length(pgs_white)` scores for the white participants and `r length(pgs_black)` for the black participants. PGS are rescaled to zero mean and unit variance within each ethnic group, so coefficient sizes are not directly comparable between ethnic groups, but are measured in standard deviations of the within-ethnic-group score. In all regressions using PGS, we control for ten within-ethnicity principal components of the DNA array data.

The key dependent variable is relative lifetime reproductive success (RLRS): number of children ever born, divided by mean number of children of people born in the same year. RLRS is calculated pooling ethnicities, i.e. treating them as members of the same biological population. The mean number of children of people born in the same year was calculated using sampling weights.

The HRS contains weights which match survey respondents to the US population. We use weights for the biomarker subsample (\*BIOWGTR in the HRS tracker file). Since half the sample enters the extended interview including biomarker data in each biannual survey, we weight individuals by either their 2010 weight or their 2012 weight. This maximizes the available sample of both black and white respondents, and should approximately match the US population of the sample cohorts between 2010 and 2012. Statistical tests are adjusted for clustering and stratification using the R "survey" package \[\@lumleysurvey2023\].

@@

tidy_pgs <- mod_pgs <- list()
tidy_pgs_ed <- mod_pgs_edyrs <- list()
tidy_pgs_sibs <- mod_pgs_sibs <- list()
tidy_pgs_gkids <- mod_pgs_gkids <- list()
tidy_pgs_x_ed <- mod_pgs_x_ed <- list()
tidy_pgs_x_inc <- mod_pgs_x_inc <- list()
tidy_pgs_x_mar <- mod_pgs_x_mar <- list()
tidy_pgs_x_agefb <- mod_pgs_x_agefb <- list()

coefs_fert_edu <- cors_fert_edu <- tidy_pgs_joined <- list()
cors_sibs_edu <- list()
cors_gkids_edu <- list()
tidy_fert_edu_phen <- list()

for (eth in eths) {

  if (eth == "white") {
    pgs <- pgs_white
  } else {
    pgs <- pgs_black
  }

  r_eth <- rand |> subset(ethnicity == eth)
  r_parents_eth <- r_parents |> subset(ethnicity == eth)
  r_child_eth <- r_child |> subset(ethnicity == eth)

@@

    form_pgs_gkids <- map(pgs, 
                       \(x) reformulate(c(x, pcs), response = "rlrs_div_h10gkid"))
  mod_pgs_gkids[[eth]] <- map(form_pgs_gkids,
                                    \(x) svyglm_quiet(x, r_child_eth))

  tidy_pgs_gkids[[eth]] <- mod_pgs_gkids[[eth]] |> 
    map(broom::tidy, conf.int = TRUE, conf.level = 0.95) |>
    list_rbind() |> 
    filter(term %in% pgs)

  cors_gkids_edu[[eth]] <- cor(tidy_pgs_gkids[[eth]]$estimate, 
                     tidy_pgs_ed[[eth]]$estimate)

@@

  if (eth == "white") {
    pgs <- pgs_white
  } else {
    pgs <- pgs_black
  }

@@

# TE we don't think it makes sense to include black estimates in the calculation. Since we know PGS in blacks are extremely weak
# and the tests overlap near 1:1 with the polygenic scores in whites
# Could just report the results for whites? 
n_tests_white <- nrow(tidy_pgs$white)
n_sig_bonf_white <- sum(tidy_pgs$white$p.value < 0.05/n_tests)

n_tests_black <- nrow(tidy_pgs$black)
n_sig_bonf_black <- sum(tidy_pgs$black$p.value < 0.05/n_tests)

@@

<!--# TE edited because 3 scores are bonferroni significant in whites. As of writing the same same three scores were signficiant regardlesss of whether we pool the blacks and whites in the bonferroni calculation.-->

We estimated coefficients of PGS on RLRS among black and white respondents separately. These are not meant to identify causal effects; recall that natural selection involves correlation, not necessarily causation, between selected characteristics and fertility. Appendix \@fig-rlrs shows coefficients for white respondents only; power is too low for individual PGS estimates to be informative in the black sample. Standard errors are large because of the relatively low sample sizes. `r n_sig_bonf` scores are significant at Bonferroni-corrected p \< 0.05/`r n_tests`. The scores are age at first birth, educational attainment and attention deficit disorder. But we are most concerned with looking at patterns across scores rather than judging the significance of individual scores.

@@

cors_fert_edu <- cors_sibs_edu <- list()
for (eth in eths) {

  if (eth == "white") {
    pgs <- pgs_white
  } else {
    pgs <- pgs_black
  }

  rand_boot <- rand |> 
    subset(ethnicity == eth) |> 
    as.svrepdesign(type = "bootstrap", replicates = n_reps)
  r_parents_boot <- r_parents |> 
    subset(ethnicity == eth) |> 
    as.svrepdesign(type = "bootstrap", replicates = n_reps)
  r_child_boot <- r_child |> 
    subset(ethnicity == eth) |> 
    as.svrepdesign(type = "bootstrap", replicates = n_reps)

  cors_fert_edu[[eth]] <- 
    withReplicates(rand_boot, theta = calc_cors_edu, var = "rlrs")
  cors_sibs_edu[[eth]] <- 
    withReplicates(r_parents_boot, theta = calc_cors_edu, var = "r10sibs")
  cors_gkids_edu[[eth]] <- 
    withReplicates(r_child_boot, theta = calc_cors_edu, var = "rlrs_div_h10gkid")
}

@@

ci_boot_fert <- map(cors_fert_edu, \(x) confint(x)[1,])
ci_boot_sibs <- map(cors_sibs_edu, \(x) confint(x)[1,])
ci_boot_gkids <- map(cors_gkids_edu, \(x) confint(x)[1,])

ci_boot_fert_lm <- map(cors_fert_edu, \(x) confint(x)[2,])
ci_boot_sibs_lm <- map(cors_sibs_edu, \(x) confint(x)[2,])
ci_boot_gkids_lm <- map(cors_gkids_edu, \(x) confint(x)[2,])
